EuPPollNet: A European database of plant-pollinator networks

### 2024-06-05 Manuscript/EuPPollNet_Ms.qmd
edits

dates1 = dates %>% 
group_by(Study_id, Network_id, Sampling_method) %>% 
summarise(Min_date = min(Date), 
          Max_date= max(Date), 
          Sampling_days = n_distinct(Date)) %>% 
mutate(Sampling_period = as.numeric(Max_date - Min_date)) %>% 
mutate(Study_number = as.integer(str_extract(Study_id, "[^_]+")), 
       .before=Study_id) %>% 
arrange(Study_number) %>% 
mutate(Sampling_period = case_when(!is.na(Sampling_period) ~ Sampling_period + 1,
TRUE ~ NA)) %>% 
mutate(Sampling_days = case_when(!is.na(Sampling_period) ~ Sampling_days,
TRUE ~ NA))
#Extract min and max number of sampling days
min_sampling_day = dates1 %>% pull(Sampling_days) %>% min(na.rm = TRUE)
max_sampling_day = dates1 %>% pull(Sampling_days) %>% max(na.rm = TRUE)
#Extract average of sampling days across networks and studies
average_sampling_days = dates1 %>% 
ungroup() %>% 
group_by(Study_id) %>% 
summarise(mean_sampling_days = mean(Sampling_days, na.rm = TRUE)) %>% 
ungroup() %>% 
summarise(mean(mean_sampling_days, na.rm = TRUE)) %>% 
pull()
#Check number of distinct sampling sites
n_sites = data %>% 
select(Latitude, Longitude) %>% 
distinct() %>% 
nrow(.)
#Check number of distinct bioregions
n_bioregion = data %>% 
select(Bioregion) %>% 
distinct()  %>% 
nrow(.)
#Read author list 
authors = readr::read_csv("../Data/Manuscript_info/Authorship_ordered.csv")
#Print authors
authors = authors %>% 
summarise(col = paste(Coauthor_name_aff, collapse="| ")) %>% 
pull()
#Read affiliation list 
aff_list = readr::read_csv("../Data/Manuscript_info/Affiliation_list.csv")
aff_list = aff_list %>% 
summarise(col = paste0(Aff_number_Aff, "\\",sep="")) %>% 
pull()

@@

**Motivation:** Pollinators play a crucial role in maintaining Earth’s terrestrial biodiversity and human food production by mediating sexual reproduction for most flowering plants. Indeed, the network of interactions formed by plants and pollinators constitutes the backbone of plant-pollinator community stability and functioning. However, rapid human-induced environmental changes are compromising the long-term persistence of plant-pollinator interaction networks. One of the major challenges for pollinator conservation is the lack of robust generalisable data capturing how plant-pollinator communities are structured across space and time. Here, we present the EuPPollNet database, a fully open and reproducible European-level database containing harmonized taxonomic data on plant-pollinator interactions referenced in both space and time, along with other ecological variables of interest. This database offers an open workflow that allows researchers to track data-curation decisions and edit them according to their preferences. Furthermore, this work assesses the taxonomic and sampling coverage of the database at the European level, and the key structural properties in plant-pollinator networks. We hope this database can facilitate future research aimed to fill in taxonomic, ecological, and geographical data gaps on plant-pollinator interactions, and explore the impacts of global change on plant-pollinator networks to guide future conservation planning for both plant and pollinator species.

**Main Types of Variables Included:** EuPPollNet contains `r prettyNum(interaction_number, big.mark=",")` interactions between plants and pollinators from `r prettyNum(n_sites, big.mark=",")` distinct locations (i.e., plant-pollinator networks), which belong to `r study_number` different studies distributed across `r country_number` European countries. In addition, information about sampling methodology, habitat type, bio-climatic region, and further taxonomic rank information for both plant and pollinator species are also provided.

@@

**Time period and grain:** Species interaction data was recorded between `r min_year` and `r max_year`. All records are time-referenced and most of the studies documented interactions in a single flowering season (`r format(round(one_year, 2), nsmall = 2)`%).

**Major taxa and level of measurement:** The database contains interaction data at the species level for `r format(round(species_level_int, 2), nsmall = 2)`% of the records, including a total of `r prettyNum(plant_spp_number, big.mark=",")` plant and `r prettyNum(pollinator_spp_number, big.mark=",")` pollinator species. The database covers `r format(round(plant_coverage, 2), nsmall = 2)`% of the European species of flowering plants, `r format(round(bee_coverage, 2), nsmall = 2)`% of bees, `r format(round(lepidoptera_coverage, 2), nsmall = 2)`% of butterflies, and `r format(round(syrphid_coverage, 2), nsmall = 2)`% of syrphids species at the European level.

@@

@@

The interactions between different plant and pollinator species within a community form complex networks. Macro-ecological analyses of the topology of these networks have revealed common properties across them, such as truncated power-law degree distributions [@jordano1987] or nestedness [@bascompte2003]. In addition, this type of large scale analyses can help understanding landscape level processes that cannot be explored at the community level, such as ecological patterns across geographic regions [@olesen2002; @traveset2016] or environmental gradients [@ramos2010; @rech2016; @saunders2023]. Although macro-ecological approaches make significant contributions to knowledge, they tend to be rare, and are strongly influenced by the spatio-temporal availability and nature of the data [@burkle2011; @trojelsgaard2013]. For instance, plant-pollinator studies tend to differ in sampling effort and methodology which affect the structure of the resulting plant-pollinator networks [@gibson2011; @jordano2016]. Further, despite little sampling effort could capture most of the relevant functional species [@hegland2010], most plant-pollinator networks have unobserved interactions because undersampling [@olesen2011; @chacoff2012]. Thus, the lack of strong spatio-temporal coverage, along with the current intrinsic limitations of sampling plant-pollinator networks, highlight the need to keep gathering and integrating informative species interaction data to properly unravel the different ecological processes that shape plant-pollinator interactions at large scales.

Here, we present the European Plant-Pollinator Networks database (EuPPollNet), which contains harmonized information on interaction data of plants and pollinators at European level. The pollinator taxonomic groups are composed by the main orders of entomofauna that visit and pollinate flowering plants in Europe. These include insect species from the orders Hymenoptera, Diptera, Lepidoptera, and Coleoptera [@willemstein1987; @potts2015], which comprise almost the totality of recorded interactions in EuPPollNet (`r format(round(main_orders_percent, 2), nsmall = 2)`%). To understand the scope of the database, we examine the taxonomic and sampling coverage of the different plant and pollinator species at European level with the help of the most up-to-date species checklists and rarefaction analyses. In addition, we investigate how two key structural metrics of plant-pollinator networks, connectance and nestedness, change across the latitudes range of studies and bioclimatic regions. Finally, we also examine whether plant-pollinator networks fulfill the expectation of being more nested than null expectations. EuPPollNet aims to cover a wide range of taxonomic groups and habitats, while also providing other variables of interest that define the ecological context and sampling methods of the study. In addition, EuPPollNet offers a transparent and accessible workflow of its data management and species harmonization that allows the database to be reused and to expand over time. This database provides a large number of community-level networks with curated and harmonized data, distinguishing it from other currently available resources that contain plant-pollinator interactions. We expect that EuPPollNet can be used to evaluate macro-ecological processes in plant-pollinator networks, guide conservation planning, and set a baseline for global change research.

@@

The EuPPollNet database is available in both .csv and .rds formats. The .csv file contains data in long format, while the .rds file stores the data in a list structure, with networks organized within each study. The file contains a total of `r nrow(my_tibble) + 1` columns that depict where and when the plant-pollinator interaction was observed (**Table 1**). To provide a standard unit of plant-pollinator interaction across studies, each row depicts a single interaction between a plant and a pollinator species. To build a plant-pollinator network matrix at a site level, users only need to group interactions by plant and pollinator species, site and study. Information about habitat type and bioclimatic region are also provided in this file. In addition, the interaction dataset includes a column that describes the presence or absence of floral counts for each study. The flower count data is provided in a separate file. Two thirds of studies contain information on floral abundance, but the methods and units differ greatly across studies. Finally, metadata at the study level is provided, including information about the authors, digital object identifier if available, sampling time and taxonomic coverage of the main pollinator groups for each study.

@@

For pollinators: (i) we first created a checklist of species names for the most representative pollinator groups at the European level by combining the most up to date published checklists of bees and syrphids [@reverte2023] with the one of butterflies [@wiemers2018]; (ii) we compared pollinator species names against the checklist and recovered some unmatched cases with restrictive fuzzy matching by using **stringdist** package [@van2014]; (iii) we fixed unmatched records when necessary and retrieved the taxonomic information for all species from GBIF; (iv) we fixed the non-found cases in the GBIF checklist and made sure that all species names from bees, syrphids and butterflies were named according to their respective species checklists.

@@

Finally, to evaluate if the presence-absence of interaction records for bees and flowering plants follows a phylogenetic pattern within the database, we calculated its phylogenetic signal  at genus and family level, respectively. The phylogenetic signal was calculated by using the _phylosig_ function from the **phytools** package [@revell2012]. We extracted the phylogenetic information for bees from a genus level phylogeny [@hedtke2013] and processed it with the help of the packages **ape** [@paradis2019], **MCMCglmm** [@hadfield2010] and **phytools**; and for plants the phylogenetic tree was obtained from a species level plant phylogeny [@smith2018] with the help of **rtree** package [@li2023].

@@

The different sites per study were described with a habitat type by the authors. As these habitats are not standardized across studies, they were standardized with the additional help of land cover information and visual checks on current satellite imagery. For each georeferenced site, the land cover information was extracted from Corine Land Cover (CLC, version 2018) with the help of the Terra package [@hijmans2022]. Based on the habitat classification from the authors and the CLC classification, we created habitat categories that intend to summarize the diversity of habitats in the EuPPollNet database (see habitat type definition in supplementary material). These categories allow a quick comparison and understanding of the habitat types from the database. Moreover, Europe is characterized by a great variety of environmental conditions that harbor different biota. Thus, to allow authors to explore the set of studies that share similar environmental conditions and species, we assigned to each site a biogeographical region. The biogeographical regions were downloaded from the European Environment Agency (version 2016) and were matched to the different sites using a spatial joint from the **sf** package [@pebesma2018].

@@

The database contains a total of `r prettyNum(pollinator_spp_number, big.mark=",")` pollinators and `r prettyNum(plant_spp_number, big.mark=",")` plant species. The coverage of the main pollinators groups occurring in Europe is `r format(round(bee_coverage, 2), nsmall=2)`% for bees, `r format(round(syrphid_coverage, 2), nsmall=2)`% for syrphids and `r format(round(lepidoptera_coverage, 2), nsmall=2)`% for butterflies (see **Figure S1** for coverage at the family level for bees and butterflies, and at the subfamily level for syrphids). Bees constitute `r format(round(bee_interactions_percent, 2), nsmall=2)`% of the interactions of the database, and `r format(round(bee_interactions_non_apis_percent, 2), nsmall=2)`% of the interactions when excluding honey bees. Within the database, `r format(round(bee_genera_percent_coverage, 2), nsmall=2)`% of bee genera have at least one species with interaction records, and the average coverage at the bee genus level is `r format(round(average_coverage_bee_genera, 2), nsmall=2)`% (**Figure 2**). In addition, the presence or absence of interactions records for bees seems not to follow any sort of phylogenetic pattern ($\lambda$ = `r format(round(lambda_phylosig_bees, 2), nsmall=2)`; _P_ = `r format(round(p_phylosig_bees, 2), nsmall=2)`). The database coverage of all flowering plant species occurring in Europe is `r format(round(plant_coverage, 2), nsmall=2)`% (**Figure 3**), with an average coverage of `r format(round(mean_plant_fam_cov, 2), nsmall=2)`% at the plant family level. Approximately, half of the plant families have at least one species with interaction records (`r format(round(percent_plant_fam_with_data, 2), nsmall=2)`%), and the presence or absence of interaction data for the different plant species also seems not to follow a statistically relevant phylogenetic pattern ($\lambda$ = `r format(round(lambda_phylosig_plants, 2), nsmall=2)`; _P_ = `r format(round(p_phylosig_plants, 2), nsmall=2)`).

@@

The estimated sampling coverage of plant and pollinator species within the database across the different sampling sites is approximately `r round(poll_coverage)`% for both taxonomic groups. This suggests that the rarefied accumulation curves of both plant and pollinator species exhibit already a "quasi-asymptotic" growth of species richness by considering the current number of sampling sites or networks (**Figure 4a-4b**). The predicted observed species richness by doubling the sampling effort on the already sampled habitat types within the database will only increase pollinator richness by `r format(round(poll_increase3, 2), nsmall=2)`% and plant richness by `r format(round(plant_increase3, 2), nsmall=2)`%. However, the sampling coverage of interactions is `r format(round(int_coverage, 2), nsmall=2)`%, and by doubling the sampling effort the predicted number of unique interactions recorded will have approximately a twofold increase (`r format(round(int_increase3, 2), nsmall=2)`%; **Figure 4c**). Similarly, to the plant and pollinator species accumulation curves across sampling sites, when we consider the accumulated pollinator richness across sampled plant species, this curve also shows a "quasi-asymptotic" growth with a sampling coverage value of `r format(round(plant_poll_coverage, 2), nsmall=2)`%. The predicted recorded pollinator species by doubling the number of plants sampled is expected to increase by `r format(round(plant_poll_increase3, 2), nsmall=2)`% (**Figure 4d**). In addition, when we explore how common are plant and pollinator species across all sampling sites, we find that the majority of plant and pollinator species tend to be regionally specific, and only a minor portion of them is being shared across a broad range of sampling sites (**Figures 4e-4f**). For instance, most plants (`r format(round(plants_under_one_percent, 2), nsmall=2)`%) and pollinators (`r format(round(polls_under_one_percent, 2), nsmall=2)`%) are exclusively found in less than 1% of sampling sites; and the most common plant (*`r most_common_plant`*) and pollinator (*`r most_common_poll`*) species are only found in `r format(round(most_common_plant_percent, 2), nsmall=2)`% and `r format(round(most_common_poll_percent, 2), nsmall=2)`% of sampling sites, respectively.

@@

The proportion of species from the major pollinator orders within the database differed across habitats and bioclimatic regions (**Figure 5**). As expected, Hymenoptera was the main taxonomic order on the majority of habitats, exceeded only by Diptera on the habitat categories of riparian vegetation, moors and heathland, and alpine grasslands. Overall, the proportion of Lepidoptera and Coleoptera species was low across all habitats but Coleoptera showed a notably increase in sclerophyllous vegetation and beaches, dunes and sands habitat categories. Similar patterns were observed when exploring the pollinator proportions by bioclimatic region. In this particular case, Hymenoptera was the predominant order in all bioclimatic regions with Diptera taking more importance in Alpine and Atlantic regions. Lepidoptera shows low proportions across all bioclimatic regions and Coleoptera seems to be only relevant in Mediterranean regions at European level. Notably, the number of studies (**Figure 5**) and sampling sites (**Figure S2**) also differs across habitats and bioclimatic regions. The habitats sampled by a higher number of studies in the database are `r studies_by_habitat$EuPPollNet_habitat[[1]]` (`r studies_by_habitat$Habitat_count[[1]]`), `r studies_by_habitat$EuPPollNet_habitat[[2]]` (`r studies_by_habitat$Habitat_count[[2]]`) and `r studies_by_habitat$EuPPollNet_habitat[[3]]` (`r studies_by_habitat$Habitat_count[[3]]`). However, the habitats that contain a higher number of sampling sites are `r networks_by_habitat$EuPPollNet_habitat[[1]]` (`r networks_by_habitat$Habitat_count[[1]]`), `r networks_by_habitat$EuPPollNet_habitat[[2]]` (`r networks_by_habitat$Habitat_count[[2]]`) and `r networks_by_habitat$EuPPollNet_habitat[[3]]` (`r networks_by_habitat$Habitat_count[[3]]`). The bioclimatic regions with a higher number of studies are `r studies_by_bioregion$Bioregion[[1]]` (`r studies_by_bioregion$Bioregion_count[[1]]`), `r studies_by_bioregion$Bioregion[[2]]` (`r studies_by_bioregion$Bioregion_count[[2]]`) and `r studies_by_bioregion$Bioregion[[3]]` (`r studies_by_bioregion$Bioregion_count[[3]]`); and the ones that contain a higher number of sampling sites are `r networks_by_bioregion$Bioregion[[1]]` (`r networks_by_bioregion$Bioregion_count[[1]]`), `r networks_by_bioregion$Bioregion[[2]]` (`r networks_by_bioregion$Bioregion_count[[2]]`) and `r networks_by_bioregion$Bioregion[[3]]` (`r networks_by_bioregion$Bioregion_count[[3]]`).

@@

EuPPollNet offers the largest set of plant-pollinator networks and plant-pollinator studies compiled to date at European level. The database contains `r prettyNum(plant_spp_number, big.mark=",")` plant and `r prettyNum(pollinator_spp_number, big.mark=",")` pollinator species with over a million interaction records. While the overall sampling coverage of species and interactions is relatively high across the sampled sites, the taxonomic coverage of plants and the main pollinator groups at the European level is still relatively low (i.e., `r round(plant_coverage, 1)`% for flowering plants and `r round(bee_coverage, 1)`% for bee species). This likely reflects the fact that most plant and pollinator species are rare, and geographically restricted. For example given that most of the plant-pollinator networks from the database are sampled on intensive grasslands, and habitat heterogeneity is a crucial factor in understanding pollinator diversity at European level [@kleijn2015; @hass2018; @martinez2022], adding studies on other habitat types is likely to result in a rapid increase of the coverage of plant and pollinator species and their interactions. Indeed, plant and pollinator species were rarely shared across multiple sites, indicating that there are few "common" species and many "rare" ones at the metaweb or continental level. This high number of "rare" species results in an upward slope of the species or interaction accumulation curves [@thompson2003]. In other words, minimal sampling efforts are capturing a substantial number of species and interactions, but achieving a comprehensive inventory will require numerous sampling events within and across habitats, particularly for plant-pollinator interactions.

>>>>

dates1 = dates %>% 
group_by(Study_id, Network_id, Sampling_method) %>% 
summarise(Min_date = min(Date), 
          Max_date= max(Date), 
          Sampling_days = n_distinct(Date)) %>% 
mutate(Sampling_period = as.numeric(Max_date - Min_date)) %>% 
mutate(Study_number = as.integer(str_extract(Study_id, "[^_]+")), 
       .before=Study_id) %>% 
arrange(Study_number) %>% 
mutate(Sampling_period = case_when(!is.na(Sampling_period) ~ Sampling_period + 1,
TRUE ~ NA)) %>% 
mutate(Sampling_days = case_when(!is.na(Sampling_period) ~ Sampling_days,
TRUE ~ NA))
#Extract min and max number of sampling days
min_sampling_day = dates1 %>% pull(Sampling_days) %>% min(na.rm = TRUE)
max_sampling_day = dates1 %>% pull(Sampling_days) %>% max(na.rm = TRUE)
#Extract average of sampling days across networks and studies
average_sampling_days = dates1 %>% 
ungroup() %>% 
group_by(Study_id) %>% 
summarise(mean_sampling_days = mean(Sampling_days, na.rm = TRUE)) %>% 
ungroup() %>% 
summarise(mean(mean_sampling_days, na.rm = TRUE)) %>% 
pull()

#Check number of distinct networks
n_networks = data %>% 
mutate(Network_id = paste0(Network_id,"_", Year)) %>% 
distinct(Network_id) %>% 
nrow(.)

#Check number of distinct sampling sites
n_sites = data %>% 
select(Latitude, Longitude) %>% 
distinct() %>% 
nrow(.)
#Check number of distinct bioregions
n_bioregion = data %>% 
select(Bioregion) %>% 
distinct()  %>% 
nrow(.)
#Read author list 
authors = readr::read_csv("../Data/Manuscript_info/Authorship_ordered.csv")
#Print authors
authors = authors %>% 
summarise(col = paste(Coauthor_name_aff, collapse="| ")) %>% 
pull()
#Read affiliation list 
aff_list = readr::read_csv("../Data/Manuscript_info/Affiliation_list.csv")
aff_list = aff_list %>% 
summarise(col = paste0(Aff_number_Aff, "\\",sep="")) %>% 
pull()

@@

**Motivation:** Pollinators play a crucial role in maintaining Earth’s terrestrial biodiversity and human food production by mediating sexual reproduction for most flowering plants. Indeed, the network of interactions formed by plants and pollinators constitutes the backbone of plant-pollinator community stability and functioning. However, rapid human-induced environmental changes are compromising the long-term persistence of plant-pollinator interaction networks. One of the major challenges for pollinator conservation is the lack of robust generalisable data capturing how plant-pollinator communities are structured across space and time. Here, we present the EuPPollNet database, a fully open and reproducible European-level database containing harmonized taxonomic data on plant-pollinator interactions referenced in both space and time, along with other ecological variables of interest. This database offers an open workflow that allows researchers to track data-curation decisions and edit them according to their preferences. Furthermore, this work assesses the taxonomic and sampling coverage of the database at the European level, and the key structural properties in plant-pollinator networks. We hope this database will facilitate future research that fills taxonomic, ecological, and geographical data gaps on plant-pollinator interactions. Also, we believe it will help evaluate patterns and drivers of plant-pollinator network change and guide future conservation planning for both plant and pollinator species.

**Main Types of Variables Included:** EuPPollNet contains `r prettyNum(interaction_number, big.mark=",")` interactions between plants and pollinators from `r prettyNum(n_networks, big.mark=",")` distinct networks (i.e., distinct sampling event in space or time), which belong to `r study_number` different studies distributed across `r country_number` European countries. In addition, information about sampling methodology, habitat type, bio-climatic region, and further taxonomic rank information for both plant and pollinator species are also provided (i.e., family, order and phylum).

@@

**Time period and grain:** Species interaction data was collected between `r min_year` and `r max_year`. All records are time-referenced and most of the studies documented interactions within a single flowering season (`r format(round(one_year, 2), nsmall = 2)`%).

**Major taxa and level of measurement:** The database contains interaction data at the species level for `r format(round(species_level_int, 2), nsmall = 2)`% of the records, including a total of `r prettyNum(plant_spp_number, big.mark=",")` plant and `r prettyNum(pollinator_spp_number, big.mark=",")` pollinator species. The database covers `r format(round(plant_coverage, 2), nsmall = 2)`% of the European species of flowering plants, `r format(round(bee_coverage, 2), nsmall = 2)`% of bees, `r format(round(lepidoptera_coverage, 2), nsmall = 2)`% of butterflies, and `r format(round(syrphid_coverage, 2), nsmall = 2)`% of syrphid species at the European level.

@@

The interactions between different plant and pollinator species within a community form complex networks. Macro-ecological analyses of the topology of these networks have revealed common properties across them, such as truncated power-law degree distributions [@jordano1987] or nestedness [@bascompte2003]. In addition, this type of large scale analyses can help understanding landscape level processes that cannot be explored at the community level, such as ecological patterns across geographic regions [@olesen2002; @traveset2016] or environmental gradients [@ramos2010; @rech2016; @saunders2023]. Although macro-ecological approaches make significant contributions to knowledge, they tend to be rare, and are strongly influenced by the spatio-temporal availability and nature of the data [@burkle2011; @trojelsgaard2013]. For instance, plant-pollinator studies tend to differ in sampling effort and methodology which affect the structure of the resulting plant-pollinator networks [@gibson2011; @jordano2016; @schwarz2020]. Most plant-pollinator networks have unobserved interactions [@olesen2011; @chacoff2012], and thus research that attempts to synthesize across published studies must have access to raw data on interactions in order to statistically account for sampling effort and completeness. Hence, this emphasizes the importance of evaluating the spatio-temporal coverage of interactions to guide future sampling efforts and accurately understand the processes that shape plant-pollinator interactions.

@@

Here, we present the European Plant-Pollinator Networks database (EuPPollNet), which contains harmonized information on interaction data of plants and pollinators at European level. The pollinator taxonomic groups are composed by the main orders of entomofauna that visit and pollinate flowering plants in Europe. These include insect species from the orders Hymenoptera, Diptera, Lepidoptera, and Coleoptera [@willemstein1987; @potts2015], which comprise almost the totality of recorded interactions in EuPPollNet (`r format(round(main_orders_percent, 2), nsmall = 2)`%). To understand the scope of the database, we examine the taxonomic and sampling coverage of the different plant and pollinator species at European level with the help of the most up-to-date species checklists and rarefaction analyses. In addition, we investigate how two key structural metrics of plant-pollinator networks, connectance and nestedness, change across the latitudes range of studies and bioclimatic regions. Given that this database contains one of the largest sets of plant-pollinator networks collated to date, and considering that the prevalence of nestedness has been evaluated only with a relatively small number of networks [@bascompte2003], we also examine whether plant-pollinator networks are more nested than expected by chance. EuPPollNet aims to cover a wide range of taxonomic groups and habitats, while also providing other variables of interest that define the ecological context and sampling methods of the study. In addition, EuPPollNet offers a transparent and accessible workflow of its data management and species harmonization that allows the database to be reused and to expand over time. This database provides a large number of community-level networks with curated and harmonized data, distinguishing it from other currently available resources that contain plant-pollinator interactions. We expect that EuPPollNet can be used to evaluate macro-ecological processes in plant-pollinator networks, guide conservation planning, and set a baseline for global change research.

@@

The EuPPollNet database is available in both .csv and .rds formats. The .csv file contains data in long format, while the .rds file stores the data in a list structure, with networks organized within each study. The file contains a total of `r nrow(my_tibble) + 1` columns that depict where and when the plant-pollinator interaction was observed (**Table 1**). To provide a standard unit of plant-pollinator interaction across studies, each row depicts a single interaction between a plant and a pollinator species. To build a plant-pollinator network matrix within a single flowering season at a site level, users only need to group interactions by plant and pollinator species, site, study and year. Information about habitat type and bioclimatic region are also provided in this file. In addition, the interaction dataset includes a column that describes the presence or absence of floral counts for each study. The flower count data is provided in a separate file. Two thirds of studies contain information on floral abundance, but the methods and units differ greatly across studies. Finally, metadata at the study level is provided, including information about the authors, digital object identifier if available, sampling time and taxonomic coverage of the main pollinator groups for each study.

@@

For pollinators: (i) we first created a checklist of species names for the most representative pollinator groups at the European level by combining the most up to date published checklists of bees and syrphids [@reverte2023] with the one of butterflies [@wiemers2018]; (ii) we compared pollinator species names against the checklist and recovered some unmatched cases with restrictive fuzzy matching by using **stringdist** package [@van2014]; (iii) we fixed unmatched records when necessary and retrieved the taxonomic information for all species from GBIF; (iv) we fixed the non-found cases in the GBIF checklist and made sure that all species names from bees, syrphids and butterflies were named according to their respective species checklists. Coleoptera species names were only checked against the GBIF checklist.

@@

Finally, to evaluate if the presence-absence of interaction records for bees and flowering plants follows a phylogenetic pattern within the database, we calculated its phylogenetic signal  at genus and family level, respectively. The phylogenetic signal was calculated by using the _phylosig_ function from the **phytools** package [@revell2012]. We extracted the phylogenetic information for bees from a genus level phylogeny [@hedtke2013] and processed it using the packages **ape** [@paradis2019], **MCMCglmm** [@hadfield2010] and **phytools**. For plants, the phylogenetic tree was obtained from a species level plant phylogeny [@smith2018] with the help of the **rtree** package [@li2023].

@@

The different sites were described with a habitat type by the authors of each study. We standardized  the habitat type using the author's description and land cover information from Corine Land Cover (CLC, version 2018) extracted using the Terra package [@hijmans2022]. Based on the habitat classification from the authors and the CLC classification, we created habitat categories that intend to summarize the diversity of habitats in the EuPPollNet database (see habitat type definition in supplementary material). These categories allow a quick comparison and understanding of the habitat types from the database. Moreover, Europe is characterized by a great variety of environmental conditions that harbor different biota. Thus, to allow authors to explore the set of studies that share similar environmental conditions and species, we assigned to each site a biogeographical region. The biogeographical regions were downloaded from the European Environment Agency (version 2016) and were matched to the different sites using a spatial joint from the **sf** package [@pebesma2018].

@@

EuPPollNet contains a total of `r prettyNum(pollinator_spp_number, big.mark=",")` pollinators and `r prettyNum(plant_spp_number, big.mark=",")` plant species. The coverage of the main pollinators groups occurring in Europe is `r format(round(bee_coverage, 2), nsmall=2)`% for bees, `r format(round(syrphid_coverage, 2), nsmall=2)`% for syrphids and `r format(round(lepidoptera_coverage, 2), nsmall=2)`% for butterflies (see **Figure S1** for coverage at the family level for bees and butterflies, and at the subfamily level for syrphids). Bees constitute `r format(round(bee_interactions_percent, 2), nsmall=2)`% of the interactions in EuPPollNet, and `r format(round(bee_interactions_non_apis_percent, 2), nsmall=2)`% of the interactions when excluding honey bees. Within the database, `r format(round(bee_genera_percent_coverage, 2), nsmall=2)`% of bee genera have at least one species with interaction records, and the average coverage of species at the bee genus level is `r format(round(average_coverage_bee_genera, 2), nsmall=2)`% (**Figure 2**). In addition, the presence or absence of interaction records for bees does not follow a phylogenetic pattern ($\lambda$ = `r format(round(lambda_phylosig_bees, 2), nsmall=2)`; _P_ = `r format(round(p_phylosig_bees, 2), nsmall=2)`). The database coverage of all flowering plant species occurring in Europe is `r format(round(plant_coverage, 2), nsmall=2)`% (**Figure 3**), with an average coverage of `r format(round(mean_plant_fam_cov, 2), nsmall=2)`% at the plant family level. Approximately, half of the plant families have at least one species with interaction records (`r format(round(percent_plant_fam_with_data, 2), nsmall=2)`%), and the presence or absence of interaction data for the different plant species also seems not to follow a statistically relevant phylogenetic pattern ($\lambda$ = `r format(round(lambda_phylosig_plants, 2), nsmall=2)`; _P_ = `r format(round(p_phylosig_plants, 2), nsmall=2)`).

@@

The estimated sampling coverage of plant and pollinator species within EuPPollNet across the different sampling sites is approximately `r round(poll_coverage)`% for both taxonomic groups. This suggests that the rarefied accumulation curves of both plant and pollinator species exhibit already a "quasi-asymptotic" growth of species richness by considering the current number of sampling sites or networks (**Figure 4a-4b**). The predicted observed species richness by doubling the sampling effort on the already sampled habitat types within the database will only increase pollinator richness by `r format(round(poll_increase3, 2), nsmall=2)`% and plant richness by `r format(round(plant_increase3, 2), nsmall=2)`%. However, the sampling coverage of interactions is `r format(round(int_coverage, 2), nsmall=2)`%, and by doubling the sampling effort the predicted number of unique interactions recorded will have approximately a twofold increase (`r format(round(int_increase3, 2), nsmall=2)`%; **Figure 4c**). When we consider the accumulated pollinator richness across sampled plant species, this curve also shows a “quasi-asymptotic” growth with a sampling coverage value of `r format(round(plant_poll_coverage, 2), nsmall=2)`%. The predicted recorded pollinator species by doubling the number of plants sampled is expected to increase by `r format(round(plant_poll_increase3, 2), nsmall=2)`% (**Figure 4d**). We find that a small portion of plant species and pollinator species are shared across a broad range of sampling sites and that most plant (`r format(round(plants_under_one_percent, 2), nsmall=2)`%) and pollinator (`r format(round(polls_under_one_percent, 2), nsmall=2)`%) species are exclusively found in less than 1% of sampling sites (**Figures 4e-4f**). The most common plant (*`r most_common_plant`*) and pollinator (*`r most_common_poll`*) species are found in `r format(round(most_common_plant_percent, 2), nsmall=2)`% and `r format(round(most_common_poll_percent, 2), nsmall=2)`% of sampling sites, respectively.

@@

The proportion of species from the major pollinator orders within the database differed across habitats and bioclimatic regions (**Figure 5**). As expected, Hymenoptera was the main taxonomic order on the majority of habitats, exceeded only by Diptera on the habitat categories of riparian vegetation, moors and heathland, and alpine grasslands. Overall, the proportion of Lepidoptera and Coleoptera species was low across all habitats but Coleoptera showed a notably increase in sclerophyllous vegetation and beaches, dunes and sands habitat categories. Similar patterns were observed when exploring the pollinator proportions by bioclimatic region. In this particular case, Hymenoptera was the predominant order in all bioclimatic regions with Diptera taking more importance in Alpine and Atlantic regions. Lepidoptera shows low proportions across all bioclimatic regions and Coleoptera is only relevant in Mediterranean regions at European level. Notably, the number of studies (**Figure 5**) and sampling sites (**Figure S2**) also differs across habitats and bioclimatic regions. The habitats sampled by a higher number of studies in the database are `r studies_by_habitat$EuPPollNet_habitat[[1]]` (`r studies_by_habitat$Habitat_count[[1]]`), `r studies_by_habitat$EuPPollNet_habitat[[2]]` (`r studies_by_habitat$Habitat_count[[2]]`) and `r studies_by_habitat$EuPPollNet_habitat[[3]]` (`r studies_by_habitat$Habitat_count[[3]]`). However, the habitats that contain a higher number of sampling sites are `r networks_by_habitat$EuPPollNet_habitat[[1]]` (`r networks_by_habitat$Habitat_count[[1]]`), `r networks_by_habitat$EuPPollNet_habitat[[2]]` (`r networks_by_habitat$Habitat_count[[2]]`) and `r networks_by_habitat$EuPPollNet_habitat[[3]]` (`r networks_by_habitat$Habitat_count[[3]]`). The bioclimatic regions with a higher number of studies are `r studies_by_bioregion$Bioregion[[1]]` (`r studies_by_bioregion$Bioregion_count[[1]]`), `r studies_by_bioregion$Bioregion[[2]]` (`r studies_by_bioregion$Bioregion_count[[2]]`) and `r studies_by_bioregion$Bioregion[[3]]` (`r studies_by_bioregion$Bioregion_count[[3]]`); and the ones that contain a higher number of sampling sites are `r networks_by_bioregion$Bioregion[[1]]` (`r networks_by_bioregion$Bioregion_count[[1]]`), `r networks_by_bioregion$Bioregion[[2]]` (`r networks_by_bioregion$Bioregion_count[[2]]`) and `r networks_by_bioregion$Bioregion[[3]]` (`r networks_by_bioregion$Bioregion_count[[3]]`).

@@

EuPPollNet offers the largest set of plant-pollinator studies and networks compiled to date at European level. The database contains `r prettyNum(plant_spp_number, big.mark=",")` plant and `r prettyNum(pollinator_spp_number, big.mark=",")` pollinator species with over a million interaction records. While the overall sampling coverage of species and interactions is relatively high across the sampled sites, the taxonomic coverage of plants and the main pollinator groups at the European level is still relatively low (i.e., `r round(plant_coverage, 1)`% for flowering plants and `r round(bee_coverage, 1)`% for bee species). This likely reflects the fact that most plant and pollinator species are rare and geographically restricted. For example given that most of the plant-pollinator networks from the database are sampled on intensive grasslands, and habitat heterogeneity is a crucial factor in understanding pollinator diversity at European level [@kleijn2015; @hass2018; @martinez2022], adding studies on other habitat types is likely to result in a rapid increase of the coverage of plant and pollinator species and their interactions. Indeed, plant and pollinator species were rarely shared across multiple sites, indicating that there are few "common" species and many "rare" ones at the metaweb or continental level. This high number of "rare" species results in an upward slope of the species or interaction accumulation curves [@thompson2003]. In other words, minimal sampling efforts are capturing a substantial number of species and interactions, but achieving a comprehensive inventory will require numerous sampling events within and across habitats, particularly for plant-pollinator interactions.
